---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# smoothqr

<!-- badges: start -->
[![R-CMD-check](https://github.com/dajmcdon/smoothqr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/dajmcdon/smoothqr/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

This package estimates multi-reponse quantile regression in the 
context that the resulting predictions should be "smooth". The typical use case would be multi-horizon time-series forecasting wherein we believe that the responses should vary smoothly.

See the related pre-print: https://arxiv.org/abs/2202.09723

## Installation

You can install the development version of smoothqr from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("dajmcdon/smoothqr")
```

## Example

This is a basic example.

```{r example}
library(smoothqr)
x <- -99:99 / 100 * 2 * pi
y <- sin(x) + rnorm(length(x), sd = .1)
XY <- lagmat(y[1:(length(y) - 20)], c(-20:20))
Y <- XY[ ,1:20]
X <- XY[ ,21:ncol(XY)]
tt <- smooth_qr(X, Y, c(.2, .5, .8), aheads = 20:1)

pl <- predict(tt, newdata = X[max(which(complete.cases(X))), , drop = FALSE])
pll <- dplyr::bind_rows(lapply(pl, tibble::as_tibble), 
                        .id = "ahead")
plot(x, y, pch = 16, col = "lightgrey")
curve(sin(x), from = -2 * pi, to = 2 * pi, add = TRUE)
abline(v = x[length(y) - 20])
pll$x <- rev(x[length(y) - 19:0])
lines(pll$x, pll$`tau = 0.5`, col = 4)
lines(pll$x, pll$`tau = 0.2`, col = 2, lty = 2)
lines(pll$x, pll$`tau = 0.8`, col = 2, lty = 2)
```

