---
title: "smoothqr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{smoothqr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r setup, message = FALSE}
library(smoothqr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
```

## Data

This vignette demonstrates `smooth_qr()` on the `covid_case_death_rates` data
from the [`epidatasets`](https://cmu-delphi.github.io/epidatasets/) package.
The data contain daily COVID-19 case and death rates (per 100,000 people) for
56 US states/territories.

```{r data}
`%nin%` <- Negate(`%in%`)
covid_case_death_rates <- covid_case_death_rates |>
  filter(geo_value %nin% c("as", "gu", "mp", "vi")) |>
  select(geo_value, time_value, case_rate, death_rate)
range(covid_case_death_rates$time_value)
n_distinct(covid_case_death_rates$geo_value) # includes 50 states, DC, and PR
```

We'll forecast future death rates using past case and death rates as
predictors, treating `2021-08-01` as the forecast date -- the date on which
the forecaster is standing and producing predictions for future days.

```{r forecast-date}
forecast_date <- ymd("2021-08-01")
```

## Building features

`smooth_qr()` expects a predictor matrix `x` and a response matrix `y`, where
each column of `y` corresponds to a different forecast horizon ("ahead"). We
use [lagmat()] to build lagged predictors (case and death rates observed 0,
7, and 14 days before the current time) and led responses (death rates 7,
14, 21, and 28 days ahead), separately for each location, then stack the
results.

```{r build-design}
aheads <- 1:28 # days ahead to forecast death_rate
lags <- c(0, 1, 2, 7, 14) # lags (days back) used as predictors

# lagmat() must be called with the full set of shifts at once so that the
# lag and lead columns stay aligned to the same set of rows.
all_shifts <- c(-aheads, lags)
mld <- max(aheads)
mlg <- max(lags)

build_design <- function(df) {
  df <- df |> arrange(time_value)
  case_full <- lagmat(df$case_rate, all_shifts)
  death_full <- lagmat(df$death_rate, all_shifts)
  colnames(case_full) <- paste0("case_", colnames(case_full))
  colnames(death_full) <- paste0("death_", colnames(death_full))
  time_padded <- c(rep(as.Date(NA), mld), df$time_value, rep(as.Date(NA), mlg))
  bind_cols(
    time_value = time_padded,
    as_tibble(case_full),
    as_tibble(death_full)
  )
}

design <- covid_case_death_rates |>
  as_tibble() |>
  group_by(geo_value) |>
  group_modify(~ build_design(.x)) |>
  ungroup() |>
  filter(!is.na(time_value)) |>
  select(!starts_with("case_ahead"))

design
```

## Fitting the model

We fit on all data up to 30 days before the forecast date, using complete cases
only (`smooth_qr()` drops incomplete rows automatically).

```{r fit}
train <- design |>
  filter(between(time_value, forecast_date - 30, forecast_date))

x_train <- train |> select(contains("lag")) |> as.matrix()
y_train <- train |> select(contains("ahead")) |> as.matrix()

fit <- smooth_qr(
  x_train,
  y_train,
  tau = c(.1, .25, .5, .75, .9),
  degree = 6L,
  aheads = aheads
)
fit
```

### Smoothed coefficients across aheads

The central idea behind `smooth_qr()` is that the predictions for
neighbouring forecast horizons should vary smoothly, rather than being
estimated independently. This is controlled by enforcing smoothness on the coefficients. We can see this by extracting the coefficients on
the response scale with `coef()` and examining how a single predictor's
coefficient evolves across aheads.

```{r coef-smoothness}
cc <- coef(fit, type = "response")

coef_df <- bind_rows(
  lapply(names(cc), function(nm) {
    as_tibble(cc[[nm]], rownames = "predictor") |> mutate(response = nm)
  })
) |>
  mutate(ahead = as.numeric(gsub("death_ahead", "", response)))

coef_df |>
  filter(predictor %in% c("case_lag0", "death_lag0")) |>
  ggplot(aes(ahead, `tau = 0.5`, color = predictor)) +
  geom_line() +
  geom_point() +
  scale_colour_manual(
    values = c("case_lag0" = "steelblue", "death_lag0" = "darkorange")
  ) +
  labs(
    x = "Forecast horizon (days ahead)",
    y = "Coefficient (tau = 0.5)",
    color = "Predictor"
  )
```

## Predicting

To forecast forward from `forecast_date`, we use each location's predictor
values observed as of that date.

```{r newdata}
newdata <- design |>
  filter(time_value == forecast_date) |>
  select(geo_value, all_of(predictor_cols))

preds <- predict(fit, newdata = newdata |> select(-geo_value))

preds_tbl <- bind_rows(
  lapply(names(preds), function(nm) {
    as_tibble(preds[[nm]]) |>
      mutate(geo_value = newdata$geo_value, response = nm)
  })
) |>
  pivot_longer(starts_with("tau"), names_to = "tau", values_to = "value") |>
  mutate(
    ahead = as.numeric(gsub("death_ahead", "", response)),
    target_date = forecast_date + ahead,
    tau = as.numeric(gsub("tau = ", "", tau))
  )

preds_tbl
```

## Visualizing a few locations

We compare the forecast distributions for California, New York, Pennsylvania,
and Utah against the death rates that were subsequently observed.

```{r plot-ca}
geo_focus <- c("ca", "ny", "pa", "ut")

actual <- covid_case_death_rates |>
  as_tibble() |>
  filter(
    geo_value %in% geo_focus,
    time_value >= forecast_date - 60,
    time_value <= forecast_date + 35
  )

pred_wide <- preds_tbl |>
  filter(geo_value %in% geo_focus) |>
  pivot_wider(
    id_cols = c(geo_value, response, ahead, target_date),
    names_from = tau,
    values_from = value,
    names_prefix = "q"
  )

ggplot() +
  geom_line(
    data = actual |> filter(time_value <= forecast_date),
    aes(time_value, death_rate)
  ) +
  geom_line(
    data = actual |> filter(time_value > forecast_date),
    aes(time_value, death_rate),
    colour = "grey60"
  ) +
  geom_ribbon(
    data = pred_wide,
    aes(x = target_date, ymin = q0.1, ymax = q0.9),
    alpha = 0.3,
    fill = "steelblue"
  ) +
  geom_ribbon(
    data = pred_wide,
    aes(x = target_date, ymin = q0.25, ymax = q0.75),
    alpha = 0.5,
    fill = "steelblue"
  ) +
  geom_line(data = pred_wide, aes(target_date, q0.5), color = "darkorange") +
  facet_wrap(~geo_value, scales = "free_y") +
  geom_vline(xintercept = forecast_date, linetype = 2) +
  labs(
    x = "Date",
    y = "Death rate (per 100k)"
  ) +
  theme_bw()
```

Note that observed death rate in California just after the forecast date shows
a data irregularity (a brief dip below zero), which is a known reporting
artifact in this dataset rather than a reflection of real negative deaths.
